### 2017-11-01 Draft2/Appending.Dynamic.tex
More cleanup

	By assumption (4), $\VarCapacitySeq_{\VarUseful + 2} \geq n > \VarCapacitySeq_{\VarUseful + 1}$. Supposing $n > \VarInitCapacity$, from (4) again I know that the rest of the sequence contains at least one term. Then $\VarUseful \geq 1$, so $\VarCapacitySeq_{\VarUseful + 2} = \VarGrowthFactor^\VarUseful\VarInitCapacity$ and $\VarCapacitySeq_{\VarUseful + 1} = \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity$. I can now derive $\VarUseful$:
	\begin{align*}
	\VarGrowthFactor^\VarUseful\VarInitCapacity &\geq n > \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
	\VarGrowthFactor^\VarUseful &\geq \frac{n}{\VarInitCapacity} > \VarGrowthFactor^{\VarUseful - 1}\\
	\VarUseful &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > \VarUseful - 1\\
	\end{align*}
	Since $\VarUseful$ is an integer,
	\begin{align*}
	\VarUseful &= \big\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \big\rceil
	\end{align*}
	If $n \leq \VarInitCapacity$, then necessarily $\VarCapacitySeq_{\VarUseful + 2} = \VarGrowthFactor^\VarUseful\VarInitCapacity$ is empty and $\VarUseful = 0$. Since $\VarUseful \geq 1$ for the previous case, we may obtain a general formula for $\VarUseful$ like so:
	\begin{align*}
	\VarUseful &= \max(\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil, 0)
	\end{align*}
	The final term in the sequence is $\VarGrowthFactor^\VarUseful\VarInitCapacity = \VarGrowthFactor^{\max(\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil, 0)}\VarInitCapacity$, completing the proof.
\end{proof}

[make lemma]
Now, consider that to grow by $\VarGrowthFactor$, $\FuncGrow$ must be called. So $\VarGrowthFactor^i\VarInitCapacity$ being present in $S_C$ means $\FuncGrow$ was called at size $\VarGrowthFactor^{i - 1}\VarInitCapacity$ for $i \geq 1$. Then $\FuncGrow$ was called at the sizes
\begin{align*}
\VarGrowSeq = 0,\ \VarInitCapacity,\ \VarGrowthFactor\VarInitCapacity,\ \VarGrowthFactor^2\VarInitCapacity,\ \ldots\ \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity
\end{align*}
Each time $\FuncGrow$ is called, $s$ items are copied where $s$ is the current size of the dynamic array. Then the total number of items copied for $n$ appends is
\begin{align*}
\sum_{s \in \VarGrowSeq} s &= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for every item in $\FuncAppend$, I finally find an explicit formula for $\FuncWrites(n)$.
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
This expression looks clean at a glance, but $\VarUseful$ implicitly depends on $n$ and uses the ceiling function and logarithms. To make it easier to analyze, I will asymptotically bound this expression.

[ref2]
First I note that $\log_{\VarGrowthFactor} k \leq \lceil \log_{\VarGrowthFactor} k \rceil < \left( \log_{\VarGrowthFactor} k \right) + 1$ for any $k > 0$, so $k \leq \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k \rceil} < \VarGrowthFactor{k}$. Since $\VarUseful = \left\lceil \log_{\VarGrowthFactor} \frac{n}{\VarInitCapacity} \right\rceil$, $\frac n \VarInitCapacity \leq \VarGrowthFactor^\VarUseful < \frac {\VarGrowthFactor{n}} \VarInitCapacity$. Then
\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \RelLessEqual \FuncWrites(n) &\RelLess n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \RelLessEqual \FuncWrites(n) &\RelLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n
\end{align*}

\HdrSpaceComplex

@@

From \ref{lem:CapacitySeq}, I know the capacity sequence of $\VarList$. Dynamic arrays also have the property that $c \in \VarCapacitySeq$ iff an array of length $c$ was allocated. Then the total length of arrays allocated is

\begin{align*}
\FuncSpace(n) &= \sum_{c \in \VarCapacitySeq} c\\
&= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^\VarUseful\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

Using the identity in [ref2] again, I asymptotically bound $\FuncSpace(n)$:

>>>>

	By assumption (4), $\VarCapacitySeq_{\VarUseful + 2} \geq n > \VarCapacitySeq_{\VarUseful + 1}$. Supposing $n > \VarInitCapacity$, from (4) again I know that the rest of the sequence contains at least one term. Then $\VarUseful \geq 1$, so $\VarCapacitySeq_{\VarUseful + 2} = \VarGrowthFactor^\VarUseful\VarInitCapacity$ and $\VarCapacitySeq_{\VarUseful + 1} = \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity$. I can now derive $\VarUseful$:
	\begin{align*}
	\VarGrowthFactor^\VarUseful\VarInitCapacity &\geq n > \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
	\VarGrowthFactor^\VarUseful &\geq \frac{n}{\VarInitCapacity} > \VarGrowthFactor^{\VarUseful - 1}\\
	\VarUseful &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > \VarUseful - 1\\
	\end{align*}
	Since $\VarUseful$ is an integer,
	\begin{align*}
	\VarUseful &= \big\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \big\rceil
	\end{align*}
	If $n \leq \VarInitCapacity$, then necessarily $\left( \VarCapacitySeq_i \right)_{i = 3}^{\VarUseful + 2}$ is empty and $\VarUseful = 0$. Since $\VarUseful \geq 1$ for the previous case, and (for this case) if $n \leq \VarInitCapacity$ then $\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil \leq 0$, a general formula for $\VarUseful$ is as follows:
	\begin{align*}
	\VarUseful &= \max(\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil, 0)
	\end{align*}
	The final term in the sequence is $\VarGrowthFactor^\VarUseful\VarInitCapacity = \VarGrowthFactor^{\max(\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil, 0)}\VarInitCapacity$, completing the proof.
\end{proof}

[make corollary]
Clearly, if $C_i$ is defined for $i > 1$, then $\FuncGrow$ must have been called at the size $C_{i - 1}$. Then the \textbf{growth sequence}, the sequence of sizes at which $\FuncGrow$ is called for $n$ appends, must be $\VarCapacitySeq$ with the last element removed:
\begin{align*}
\VarGrowSeq = \left( \VarCapacitySeq_i \right)_{i = 1}^{|\VarCapacitySeq| - 1}
\end{align*}
Each time $\FuncGrow$ is called, $s$ items are copied where $s$ is the current size of the dynamic array. Then the total number of items copied for $n$ appends is
\begin{align*}
\sum_{s \in \VarGrowSeq} s &= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for every item in $\FuncAppend$, I finally find an explicit formula for $\FuncWrites(n)$.
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
This expression looks clean at a glance, but $\VarUseful$ implicitly depends on $n$ and uses the ceiling function and logarithms. To make it easier to analyze, I will asymptotically bound this expression.

[ref2]
First I note that $\log_{\VarGrowthFactor} k \leq \lceil \log_{\VarGrowthFactor} k \rceil < \left( \log_{\VarGrowthFactor} k \right) + 1$ for any $k > 0$, so $k \leq \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k \rceil} < \VarGrowthFactor{k}$. Since $\VarUseful = \left\lceil \log_{\VarGrowthFactor} \frac{n}{\VarInitCapacity} \right\rceil$, $\frac n \VarInitCapacity \leq \VarGrowthFactor^\VarUseful < \frac {\VarGrowthFactor{n}} \VarInitCapacity$. Then
\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \RelLessEqual \FuncWrites(n) &\RelLess n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \RelLessEqual \FuncWrites(n) &\RelLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n
\end{align*}
\HdrSpaceComplex

@@

From \ref{lem:CapacitySeq}, I know the capacity sequence of $\VarList$. Dynamic arrays also have the property that $c \in \VarCapacitySeq$ iff an array of length $c$ was allocated. Then the total length of arrays allocated is
\begin{align*}
\FuncSpace(n) &= \sum_{c \in \VarCapacitySeq} c\\
&= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^\VarUseful\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Using the identity in [ref2] again, I asymptotically bound $\FuncSpace(n)$:

### 2017-11-01 Draft2/Appending.Growth.tex
Even more
More cleanup

\begin{lemma}
	The capacity of a non-empty growth array grows by the constant factor $\VarGrowthFactorNew$.
\end{lemma}

@@

	For $k > 2$, by induction $c_i(k) = c_f(k - 1) = \VarGrowthFactorNew^{k - 2}\VarInitCapacityNew$ and $h_i(k) = h_f(k - 1) = (\VarGrowthFactorNew^{k - 2} - \VarGrowthFactorNew^{k - 3})\VarInitCapacityNew$. I wish to show $c_f(k) = \VarGrowthFactorNew^{k - 1}\VarInitCapacityNew$ and $h_f(k) = (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew$. Because $k > 2$, the algorithm will calculate $h_f$ as $\VarGrowthFactorNew$ times $h_i$. Then
	\begin{align*}
	h_f &= \VarGrowthFactorNew h_i\\
	&= \VarGrowthFactorNew (\VarGrowthFactorNew^{k - 2} - \VarGrowthFactorNew^{k - 3})\VarInitCapacityNew\\
	&= (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew
	\end{align*}
	and
	\begin{align*}
	c_f &= c_i + h_f\\
	&= \VarGrowthFactorNew^{k - 2}\VarInitCapacityNew + (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew\\
	&= \VarGrowthFactorNew^{k - 1}\VarInitCapacityNew
	\end{align*}
	as desired.
\end{proof}

@@

Now I take a look at $\FuncGrow$. Unlike dynamic arrays, the number of writes performed does not equal $i$ if the current size is $i$. No items are being copied; the only source of writes is appending a buffer to the tail. The goal is to find the number of writes that causes.

Let $\FuncWritesByGrow(n)$ denote the total number of writes made by $\FuncGrow$. From [make lemma], I know that $\FuncGrow$ is called at the sizes

\begin{align*}
S_R = 0,\ \VarInitCapacityNew,\ \VarGrowthFactorNew\VarInitCapacityNew,\ \VarGrowthFactorNew^2\VarInitCapacityNew,\ \ldots\ \VarGrowthFactorNew^{\lceil \log_{\VarGrowthFactorNew} n - \log_{\VarGrowthFactorNew} \VarInitCapacityNew \rceil - 1}\VarInitCapacityNew
\end{align*}

Let $k$ be the number of times the tail is appended to. Since the tail is a dynamic array, the number of writes it makes is $\FuncWrites(k)$. Since $\FuncGrow$ is called $|S_R|$ times and it appends to the tail each time except for the first, $k = |S_R| - 1 = \lceil \log_{\VarGrowthFactorNew} n - \log_{\VarGrowthFactorNew} \VarInitCapacityNew \rceil$. Then the formula for $\FuncWritesByGrow(n)$ is

\begin{align*}
\FuncWritesByGrow(n) = k + \left( \frac{\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k - \log_{\VarGrowthFactor} \VarInitCapacity \rceil} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

where $\VarInitCapacity$ $\VarGrowthFactor$ are the capacity of the \textit{tail} (not the growth array). I finally conclude that

\begin{align*}
\FuncWritesNew(n) = n + k + \left( \frac{\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k - \log_{\VarGrowthFactor} \VarInitCapacity \rceil} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

>>>>

\begin{lemma}
\label{lem:GrowthArraysGrowthFactor}
	The capacity of a non-empty growth array grows by the constant factor $\VarGrowthFactorNew$.
\end{lemma}

@@

	For $k > 2$, by induction $c_i(k) = c_f(k - 1) = \VarGrowthFactorNew^{k - 2}\VarInitCapacityNew$ and $h_i(k) = h_f(k - 1) = (\VarGrowthFactorNew^{k - 2} - \VarGrowthFactorNew^{k - 3})\VarInitCapacityNew$. I wish to show $c_f(k) = \VarGrowthFactorNew^{k - 1}\VarInitCapacityNew$ and $h_f(k) = (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew$. Because $k > 2$, the algorithm will calculate $h_f(k)$ as $\VarGrowthFactorNew$ times $h_i(k)$. Then
	\begin{align*}
	h_f(k) &= \VarGrowthFactorNew h_i(k)\\
	&= \VarGrowthFactorNew (\VarGrowthFactorNew^{k - 2} - \VarGrowthFactorNew^{k - 3})\VarInitCapacityNew\\
	&= (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew
	\end{align*}
	and
	\begin{align*}
	c_f(k) &= c_i(k) + h_f(k)\\
	&= \VarGrowthFactorNew^{k - 2}\VarInitCapacityNew + (\VarGrowthFactorNew^{k - 1} - \VarGrowthFactorNew^{k - 2})\VarInitCapacityNew\\
	&= \VarGrowthFactorNew^{k - 1}\VarInitCapacityNew
	\end{align*}
	as desired.
\end{proof}

Now that \ref{lem:GrowthArraysGrowthFactor} and consequently \ref{lem:CapacitySeq} are proved for growth arrays, I examine the write cost of $\FuncGrow$. Unlike dynamic arrays, $\FuncGrow$ does not make $s$ writes if the current size is $s$. In fact, no items are copied; the only source of writes is when a buffer is appended to the tail, since the tail is a dynamic array.

Let $\FuncWritesByGrowNew(n)$ denote the total number of writes made by $\FuncGrow$, and let $\VarNumItemsTail$ be the size of the tail. Consider that [make corollary] also holds true for growth arrays, so $\FuncGrow$ is called $|\VarGrowSeq|$ times. A buffer is appended to the tail every time except the first, so the size of the tail is
\begin{align*}
\VarNumItemsTail &= |\VarGrowSeq| - 1 = |\VarCapacitySeq| - 2 = \VarUsefulNew
\end{align*}
Then the formula for $\FuncWritesByGrowNew(n)$ is simply $\FuncWrites(\VarUsefulNew)$, where $\FuncWrites$ is the write cost function for dynamic arrays. Finally, adding the writes made by $\FuncAppend$, the formula for $\FuncWritesNew(n)$ is
\begin{align*}
\FuncWritesNew(n) = n + \FuncWrites(\VarUsefulNew)
\end{align*}

@@
